\documentclass[10pt]{report}

% ---------- Package Setup ----------
\usepackage[utf8]{inputenc}
\usepackage{amsmath, amssymb}
\usepackage{graphicx}
\usepackage{cite}
\usepackage{hyperref}
\usepackage{geometry}
\usepackage{titlesec}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{tocloft}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{setspace}
\usepackage{soul}


\newcommand{\rs}[1]{\noindent{\color{red}\textbf{RS:} #1}}
% \newcommand{\hl}[1]{\noindent{\color{blue}\textbf{HL:} #1}}
\newcommand{\rev}[1]{\textcolor{red}{#1}}
\newcommand{\mm}[1]{\noindent{\color{cyan}\textbf{MM:} #1}}

\setstretch{1.5}


\hypersetup{
    colorlinks=true,
    linkcolor=black,
    citecolor=black,
    urlcolor=black
}


% ---------- Page Margin ----------
\geometry{
 a4paper,
 total={6in, 9in},
 left=1.2in,
 right=1.2in,
 top=1.2in,
}


% ---------- Chapter/Section Format ----------

% Chapter title format
\titleformat{\chapter}[hang]
  {\normalfont\large\bfseries}
  {\thechapter\ } % Add a space after the number
  {0pt}
  {}

% Adjust spacing before/after chapter titles
\titlespacing*{\chapter}{0pt}{-10pt}{20pt}

% Section title format
\titleformat{\section}
  {\large\bfseries}
  {\thesection}{1em}{}

% Adjust spacing between sections in TOC
\setlength{\cftbeforesecskip}{6pt}

% ---------- TOC / List Titles ----------

% Font size of Contents / List of Figures / List of Tables
\renewcommand{\cfttoctitlefont}{\large\bfseries}
\renewcommand{\cftloftitlefont}{\large\bfseries}
\renewcommand{\cftlottitlefont}{\large\bfseries}

% Vertical spacing between list titles and top of the page
\setlength{\cftbeforetoctitleskip}{-5pt}   % For Table of Contents
\setlength{\cftbeforeloftitleskip}{-5pt}   % For List of Figures
\setlength{\cftbeforelottitleskip}{-5pt}   % For List of Tables










% --------- preamble -----------
\usepackage{amsthm} 

\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\newtheorem{estimator}[definition]{Estimator}







% ---------- Document ----------
\begin{document}
\pagenumbering{gobble}

% ---------- Title Page ----------
\begin{titlepage}
\centering
{\large \textbf{\MakeUppercase{Exploring Ensemble Robustness with CrossMax: From Provable Defenses to Confidence Analysis}}} \\[0.8cm]


\centering{\large by \\ Hao Luo} \\
Bachelor of Engineering, The Chinese University of Hong Kong, Shenzhen, 2024 \\ [2cm]

\large A thesis \\
presented to Toronto Metropolitan University \\ [0.5cm]
in partial fulfillment of the \\
requirements for the degree of \\

{\large Master of Engineering} \\
in the program of \\
Electrical, Computer, and Biomedical Engineering \\[2cm]

Toronto, Ontario, Canada, August 2025 \\[2cm]

\copyright\ Hao Luo, 2025



\end{titlepage}


\clearpage
\pagenumbering{roman}   % 前置部分用小写罗马数字
\setcounter{page}{2}    % 从(ii)开始



\begin{center}
\textbf{AUTHOR’S DECLARATION FOR ELECTRONIC SUBMISSION OF A THESIS}
\end{center}

\vspace{1em}

\indent I hereby declare that I am the sole author of this thesis. This is a true copy of the thesis, including  
any required final revisions, as accepted by my examiners.

\noindent I authorize Toronto Metropolitan University to lend this thesis to other institutions or individuals for the  
purpose of scholarly research.

\noindent I further authorize Toronto Metropolitan University to reproduce this thesis by photocopying or by other  
means, in total or in part, at the request of other institutions or individuals for the purpose of scholarly  
research.

\noindent I understand that my thesis may be made electronically available to the public.
\clearpage

\newpage

\clearpage


\begin{center}
\textbf{Abstract}
\end{center}

\begin{doublespacing}

\indent

Adversarial robustness remains a critical challenge in deep learning, where small perturbations can severely degrade model performance. Ensemble methods offer potential by leveraging model diversity, yet their robustness trade-offs are not fully understood. This project conducts an empirical study of ensemble robustness against strong white-box attacks on CIFAR-10, focusing on ensembles built from models trained via standard procedures and Linear Programming (LP).

We introduce CrossMax, a novel aggregation technique producing set-valued predictions to capture uncertainty under attack, along with a derived CrossMax Confidence measure. Under adversarial perturbations, ensembles achieve higher robust accuracy than their single-model counterparts, highlighting the benefit of model diversity under strong attacks. In particular, LP-trained ensembles, despite lower clean accuracy, consistently surpass standard-trained ones in robustness and also exhibit greater prediction diversity. Notably, CrossMax-Ambiguous ensembles achieve up to 15\% higher robust accuracy than traditional voting schemes, particularly under larger perturbation budgets. Furthermore, CrossMax Confidence reflects adversarial sensitivity more reliably than Softmax Confidence, providing a practical indicator of robustness.

This study demonstrates that combining LP training with CrossMax aggregation enhances adversarial robustness beyond single LP models, highlighting the promise of uncertainty-aware ensemble methods for advancing defense strategies.

\end{doublespacing}
\clearpage



\tableofcontents

\clearpage


\listoffigures

\clearpage

\listoftables




\include{chapters/introduction}
\include{chapters/literature_review}
\include{chapters/study_design}
\include{chapters/experiment}
\include{chapters/conclusion_future}
\include{chapters/appendix}


% ----------------- REFERENCES -----------------
\bibliographystyle{IEEEtran}
\bibliography{references}

\end{document}
